---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The `teffects` package implements several estimators for causal effects in observational data. 
They all start from the same problem and use the same underlying causal assumption: conditional independence.
They differ in how they use the observed data to form estimates.

This vignette explains that logic first. 
It then shows the common formula interface and gives a short guide to the estimator families. 
The mathematical derivations and inference are discussed in [Mathematical details](mathematical-details.html).

For treatment levels $d \in \{0, \ldots, K \}$, let $\mu_d = \mathbb{E}\{Y_i(d)\}$ be the average value of the $d$-th potential outcome for the population.
Then, (overall) average treatment effects are defined as
\[
  ATE(d) = \mu_d - \mu_0.
\]
In the case of binary treatments, this is the $\text{ATE}$.
This quantity averages over every unit in the population to form the overall average.

Sometimes, we want to know the average effect of treatment on the units that actually select into treatment. 
We define this as the average treatment effect on the treated.
For a binary treatment, the ATT is
\[
  ATT = \mathbb{E}\{ Y_i(1) - Y_i(0) \ \mid \ D_i = 1\}.
\]

In general, we have $K$ ATT parameters, one for each treatment value:
$$
  ATT(d \ \mid \  d) = \mathbb{E}\{ Y_i(1) - Y_i(0) \ \mid \ D_i = d \}.
$$

We will make two main assumptions to identify these treatment effect parameters: conditional independence and overlap.

## The problem with a difference in means

Let \(D_i\) denote treatment, \(X_i\) the pre-treatment covariates, and \(Y_i(d)\) the potential outcome under treatment level \(d\). 
We observe only

\[
  Y_i = \sum_{d} 1(D_i = d) Y_i(d).
\]

The fundamental problem of causal inference is that we only observe a single potential outcome for each unit. 
In a randomized experiment, treatment assignment is independent of potential outcomes, so the control group provides a good estimate of what would have happened to treated units had they not been treated.

Most empirical work in economics uses observational data. Units select into treatment, so treated and untreated units can differ in ways that matter for the outcome. 
For example, college graduates may have higher school smarts, more educated parents, and better high schools. 
Those differences would lead to higher earnings even without college, so a comparison of average earnings does not isolate the effect of attending college.

For a binary treatment, the difference in observed means can be written as

\[
E(Y_i \ \mid \ D_i = 1) - E(Y_i \ \mid\ D_i = 0)
  = \underbrace{\mathbb{E}\{Y_i(1) - Y_i(0) \ \mid \  D_i=1\}}_{\text{ATT}}
  + \underbrace{\mathbb{E}\{Y_i(0) \ \mid \  D_i = 1\} - \mathbb{E}\{Y_i(0) \ \mid \  D_i = 0\}}_{\text{selection bias}}.
\]

The second term is the problem and is why just comapring average outcomes is not sufficient to identify treatment effects. 
Our conditiona independence assumption will allow us to remove this term by comparing units with the same relevant observed characteristics.

## Selection on observables

The key assumption is the conditional independence assumption, also called the
selection-on-observables assumption:

\[
  ( Y_i(0), Y_i(1) )\mathrel{\perp\!\!\!\perp} D_i \ \mid \ X_i = x.
\]

Among units with the same \(X_i=x\), treatment assignment is independent of the potential outcomes. 
In effect, treatment is "as good as randomly assigned" within each covariate cell. 
This is like having many small randomized experiments, one for each value of \(x\).

There is always going to be debate about what variables are missing in \(X_i\).
As I put it in my class:

> Author: "X_i includes a lot of important factors that drive selection
> into treatment!" Reviewer: "I think there are other omitted variables that
> drive selection into treatment!!"

The credibility of a selection-on-observables design depends on that debate.

## From conditional comparisons to causal effects

Suppose first that \(X_i\) is a discrete variable such as gender. 
Under conditional independence assumption, we have two (as good as) random experiments: one for men and one for women.

The solution is to estimate a treatment effect within each group:
\[
 \tau(x) = E(Y_i \ \mid \ D_i = 1, X_i = x) - E(Y_i \ \mid \ D_i = 0, X_i =  x).
\]
That is, we will estimate conditional average treatment effects for each value of the covariate using difference in means.

Under conditional independence, this is the conditional average treatment effect,
\[
  \tau(x) = \mathbb{E} \{ Y_i(1) - Y_i(0) \ \mid \ X_i = x\}.
\]

The overall effect is an average of these conditional effects
\[
  ATE = \int \tau(x) \, dP(X = x).
\]

This approach of performing many difference-in-means estimators works well when $X_i$ is a single or a handful of variables, but quickly becomes infeasible as $X_i$ becomes larger.
In practice, trying to find people with the exact same value of $X_i$ when you have many control variables becomes too difficult to produce any good estimates.
The problem is called the **curse of dimensionality**.

Therefore, we will need to reach for alternative estimators. 
The goal of this package is to make these available to users.


## Propensity Score

Above we discuss the difficulties with controlling for a large number of $X_i$ variables.
It turns out that under conditional independence, it is sufficient to control for a single scalar variable: the propensity score.

That is, under conditional independence, we have:
\[
  (Y_i(0), Y_i(1)) \mathrel{\perp\!\!\!\perp} D_i \mid \pi(X_i).
\]

This reduces a high-dimensional comparison ($X_i$) to one treatment probability ($\pi(X_i)$).

## Overlap

In addition to the conditional independence assumption, we will need the **overlap assumption**.

The overlap assumption involves the **propensity score** which is the probability of being in the treatment group given your covariate values:
$$
  \pi(x) = P(D_i = 1 \ \mid\  X_i = x)
$$

The overlap condition requires
\[
  0 < P(D_i = 1 \ \mid\  X_i = x) < 1
\]
for all values of $x$.
Without overlap, we will not be able to find treated and untreated observations with the same value of $x$ to compare.

In practice is it common to trim observations with poor overlap (propensity score close to 0.), but it's important to remember that this changes the target population (ATE among units satisfying overlap).




## Alternative treatment effect estimators

The estimators below all are based on the conditional independence assumption and the overlap assumption.
They just differ in how they compute treatment effects under these assumptions.

Here is a high-level summary of the estimators, but we will go into details on each one below:

| Function | Adjustment strategy | Main modeling task |
|---|---|---|
| `ra()` | Regression adjustment | Fit an outcome model in each treatment arm |
| `ipwra()` | IPW regression adjustment | Fit propensity-weighted outcome models |
| `ipw()` | Inverse-probability weighting | Estimate treatment probabilities or solve balance equations |
| `aipw()` | Augmented IPW | Combine outcome and treatment models |
| `psmatch()` | Propensity-score matching | Match on estimated treatment probabilities |
| `nnmatch()` | Nearest-neighbor matching | Match directly on covariates |

For this vignette, we will focus on estimating the $\text{ATT}$, but the intuition will stand for $\text{ATE}$ parameters. 
The only practical difference is that the $\text{ATE}$ requires us to guess both the treated units' untreated potential outcome *and* the control units' treated potential outcome.
To begin, we will generate example data to use in our discussion:

```{r data}
library(teffects)

set.seed(2026)
n <- 300
x1 <- rnorm(n)
x2 <- rbinom(n, 1, 0.45)
probability <- plogis(-0.2 + 0.8 * x1 - 0.6 * x2)
d <- rbinom(n, 1, probability)
y0 <- 1 + 1.5 * x1 - 0.8 * x2 + rnorm(n)
example_data <- data.frame(y = y0 + 2 * d, d, x1, x2)
```

All estimators mark treatment with `treat()`:

```r
y ~ treat(d, ref = 0) + x1 + x2
```

The left side is the outcome. `ref` selects the comparison level. The remaining
terms describe pre-treatment adjustment covariates. Use ordinary R formula
syntax for factors, transformations, and interactions. Use `treatment_fml`
when the treatment model needs a different covariate specification from the
outcome model. Smooth estimators use `stat = "ate"` by default. Use
`stat = "pomeans"` for all potential-outcome means, `stat = "att"` for the
treated population, or `stat = "atc"` for the control population when the
estimator supports that target.


### Regression adjustment

With many covariates or continuous \(X_i\), exact stratification becomes
impractical. The curse of dimensionality means that most covariate cells have
few or no observations. Regression adjustment models the conditional outcome
functions instead.

The main issue for estimating the $\text{ATT}$ is estimating the untreated potential outcome, $Y_i(0)$.

To do so, we will try to estimate this conditional expectation function
\[
  m_0(x) = E(Y_i(0) \mid X_i = x).
\]
Since we only observe $Y_i(0)$ for untreated units, we will run a regression of $Y_i$ on $X_i$  using the untreated units, i.e. $D_i = 0$. 

Then, we will estimate the treated unit's untreated potential outcome with $\hat{Y}_i(0) = X_i \hat{\beta}_0$ where the $_0$ is used to emphasize this is the regression ran on the untreated group. 
In practice, we will want our models to be more flexible, so we can include transformations of the underling $X_i$ variables. 

For a binary treatment, the ATT estimate can be formed as
\[
 \widehat{\text{ATT}}^{RA}
  = \frac{1}{n_1} \sum_{i: D_i = 1} \{Y_i - \widehat m_0(X_i)\}.
\]

```{r ra-fit}
ra(y ~ treat(d, ref = 0) + x1 + x2, data = example_data)
```


### Matching

Matching avoids a global functional-form assumption by finding observations
that are close in \(X_i\) or in an estimated propensity score. Let
\[
  \mathcal J_M(i)
\]
be the set of (M) "closest" observations from the opposite treatment group (we will discuss how to define closest in a moment).

The missing potential outcome is imputed by their average:
$$
  \hat{Y}_{i}(0) = \sum_{j \in \mathcal{J}_M(i) Y_j}
$$. 
The treatment effect is then formed as
\[
 \widehat{\text{ATT}}^{Matching}
  = \frac{1}{n_1} \sum_{i: D_i = 1} \{Y_i - \sum_{j \in \mathcal{J}_M(i)}\}.
\]

We can either use `nnmatch` to match directly on the full set of covariates or use  `psmatch()` matches on the estimated propensity score. 

```{r matching-fit}
(ps_fit <- psmatch(y ~ treat(d, ref = 0) + x1 + x2, data = example_data, nneighbor = 1))
(nn_fit <- nnmatch(y ~ treat(d, ref = 0) + x1 + x2, data = example_data, nneighbor = 1))
```

There are a lot of options on how to do matching.
The [Matching details](stata-nnmatch.html) vignette documents options the
nearest-neighbor rules used by `teffects nnmatch`, including ties, exact
matching, calipers, and standard errors.
Both match with replacement.
More neighbors reduce variance but can make the matches less similar. 
A caliper can rule out comparisons that are too far apart. 
Exact matching in `ematch` is applied before the distance calculation, and `biasadj` can correct the remaining smooth covariate difference.

Matching makes each comparison explicit, but it does not make incomparable
units comparable. Check the distance between matches and covariate balance.

In practice, matching is not perfect and we will have some differences in the distribution of $X$ for treated and control units.
It is recommended to do a regression adjustment estimator between the treated and the matched control group, using `biasadj` argument.
```{r}
(nn_fit_bias_adj <- nnmatch(y ~ treat(d, ref = 0) + x1 + x2,
  data = example_data, nneighbor = 1,
  biasadj = ~ x1 + x2
))
```

### Propensity-score weighting 

Inverse-probability of treatment weighting then gives more weight to observations whose treatment status is less expected given their covariates. 
For the ATT, treated observations receive their usual weight and controls receive IPTW weights:
\[
  \frac{\pi(X_i)}{1 - \pi(X_i)}.
\]
Our weighting procedure puts more weight on control units who have a large propensity score because they look like they should be treated.

The treatment effect estimator is formed as:
\[
 \widehat{\text{ATT}}^{IPTW}
  = \frac{1}{n_1} \sum_{i: D_i = 1} Y_i - \frac{1}{n_0} \sum_{j: D_j = 0} \frac{\pi(X_j)}{1 - \pi(X_j)} Y_j.
\]

That is, we do a difference in means using a weighted control group. 

There are different ways we could choose to estimate the propensity score. 
`ipw()` uses normalized weighted means. The options
`psmethod = "logit"` and `"probit"` estimate treatment probabilities by
maximum likelihood. `"ipt"` and `"cbps"` choose probabilities by solving
equations that target covariate balance.

```{r weighting-fit}
(ipw_fit <- ipw(y ~ treat(d, ref = 0) + x1 + x2, data = example_data))
(ipt_fit <- ipw(y ~ treat(d, ref = 0) + x1 + x2, data = example_data, psmethod = "ipt"))
(ipt_fit <- ipw(y ~ treat(d, ref = 0) + x1 + x2, data = example_data, psmethod = "cbps"))
```

Extreme propensity scores create extreme weights (e.g. a propensity score of 0.999 gives that control unit a weight of 999).  
Therefore, it is important to inspect the fitted scores and their overlap before interpreting an IPTW estimate. 
The default behavior is to stop when a score violates `pstolerance`.
The opt-in `psaction = "clamp"` changes the propensity scores to be in some range (e.g. [0.05, 0.95]) and `psaction = "trim"` drops any units outside of a given range.


### Doubly robust estimators

The final two estimators are called "doubly robust estimators". 
Both use an outcome model \(m_d(x)\) and a treatment model \(\pi_d(x)\).

First, we have the IPWRA estimator.
`ipwra()` uses the treatment probabilities to weight the arm-specific outcome regressions. 
After running the IPW-weighted least squares regression using IPTW weights, regression adjustment proceeds as normal. 

Second, we have the augmented IPW estimator.
`aipw()` combines outcome predictions with inverse-probability-weighted residuals. 
The AIPW estimator for a potential-outcome mean is

\[
  \widehat{\mu}_d^{AIPW} = \frac{1}{n}\sum_{i=1}^n \widehat{m}_d(X_i) + 
  \frac{1}{n}\sum_{i=1}^n  \frac{1(D_i=d)}{\widehat e_d(X_i)} \{Y_i-\ \widehat{m}_d(X_i)\}.
\]

You can view it as a regression adjustment estimate for the average potential outcome with a "correction" for modelling errors: the estimator then rebalances $Y_i-\ \widehat{m}_d(X_i)$

Its double-robust property has a precise meaning: under the other identifying and regularity conditions, it remains consistent if either the outcome model or the treatment model is correctly specified. 
It does not protect against an omitted confounder, failure of overlap, or misspecification of both models.

```{r doubly-robust-fit}
(aipw_fit <- aipw(y ~ treat(d, ref = 0) + x1 + x2, data = example_data))
(ipwra_fit <- ipwra(y ~ treat(d, ref = 0) + x1 + x2, data = example_data))
```

